Environmental suitability throughout the late Quaternary explains population genetic diversity.

### 2023-08-03 manuscript_english.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels},
  pdfauthor={Norma Hernandez, Angel Robles Nathan Upham},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context. Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly. We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species. Our results reveal that in cases of genetic isolation by
geographic distance, suitability prevalence along time corresponds to the genetic fixation
index of populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower \(F_{st}\) value. This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.

@@

Current species distributions do not always correspond to their
historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This
is because environmental conditions are highly dynamic over time, and species
tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).

@@

With this approach we expect to predict the geographic pattern of population-genetic
structure from abiotic environmental information alone. This approach is strongly
driven by advances in currently available climate simulations (Leonardi
et al. 2023; Krapp et al. 2021), as well as next-generation sequencing
data, and supported by both ecological niche (Thorup et al. 2021;
Nogués-Bravo 2009) and population genetics theories (Lira-Noriega and
Manthey 2014). Importantly, these predicted patterns of population structure will serve as a type of \textit{a priori} hypothesis for the influence of abiotic factors alone, with the empirical detection of deviations from this hypothesis indicating that biotic factors such as competition or predation from other species have also impacted the observed population structure.

In this work, we propose a method to find what we call the Suitability Prevalence Area
(SPA) that is the time-averaged area of highest environmental suitability for a given species, defined as locations with ≥ 90\% suitability over a historical time period. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) predicting the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.

As a case study, we consider Tassel-eared Squirrels (\emph{Sciurus aberti}), a species of rodent that is currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution (Bono et al. 2018; Burgin et al. 2018). To obtain the SPA for \textit{S. aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.

Subsequently, we delimited historical endemic areas to locations where
the prevalence of suitability remained at 90\% during this period.

Subsequently, we calculated the fixation index (\(F_{st}\)) of the genetic relatedness among populations and compared that to the SPA index of historical endemism, constructing a statistical model of the fixation index as a function of the SPA. With this statistical model, the \(F_{st}\) values were projected to the current geographic distribution of \textit{S. aberti} to obtain a predictive map of expected \(F_{st}\) values over space.

@@

In other words, the sites in the geography \(\mathbf{G}_o\) that present
positive population growth rates are related to favorable environmental conditions. If we take
the set of places in ambient space where these values are greater than a
certain threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, as \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

@@

Locally, the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

@@

This equation was proposed by Soberon and Peterson Soberón and Peterson
(2020) to relate demographics to a function of evolutionary fitness and
environment. In that study they tested the validity with monthly
variables and empirical values on a short time scale. relating life
history parameters and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

@@

The goal of this theoretical approach is not to know the shape of the
function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)), i.e.~:

\[
N_i \propto  S_i(\vec{e}_i)
\] Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)

@@

On the other hand, loss of heterozygosity may be related to the amount
of genetic variation present in the absence of mutation and selection,
so theoretically, one would expect a strong correlation between
effective population size and heterozygosity on the basis of population
genetics. We start from the assumption that in the absence of gene flow
between subpopulations, the rate of fixation by isolation depends on the
effective population size \(N_e\) when subpopulations diverge from a
common ancestor in \(t\) generations. Wright (1943):

@@

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
We substitute the harmonic mean into the above equation, considering
that although the number of generations is not exactly equal to the time
of the fixation index, they are expected to be very close.

@@

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
where the \(F_{st}\) is lower are in areas where the prevalence over
time is higher, i.e.~environmental conditions remain suitable for a
significant period of time and therefore no effects of genetic drift of
populations have occurred. Conversely, it is expected that in
populations with a high fixation index value, suitability will not have
prevailed over the same time interval even though conditions currently
exist and populations are present.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past analogous climates and retain projections only for
  sites with climates analogous to the present.
\item
  Generate a harmonic mean suitability map for all time slices.
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  Obtain georeferenced points of genetically structured populations with
  fixation index values.
\item
  Extract values from the SPA map with these points.
\item
  A linear model of the fixation index value with respect to the SPA is
  performed.
\item
  With the linear model all SPA values of suitability are interpolated.
\item
  A map is generated with all the values interpolated by the model.
\end{itemize}

@@

\hypertarget{analogy-of-environmental-conditions}{%
\paragraph{Analogy of environmental
conditions}\label{analogy-of-environmental-conditions}}

To be certain that areas in the past are climatically analogous to the
present we implemented the mobility-orient parity (MOP) test. This
method generates a value from \(0\) to \(1\) according to the
environmental analogy of past conditions to the present to eliminate
geographic areas in the projections to the past outside of climates
analogous to the present (Owens et al. 2013).

@@

\hypertarget{fixation-index-as-a-function-of-spa}{%
\subsubsection{Fixation index as a function of
SPA}\label{fixation-index-as-a-functino-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_{st}\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, geografic references were
obtained for each population considered in the fixation index matrix.
The SPA value for each population was
extracted for each population site. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

\begin{longtable}[]{@{}ccc@{}}
\toprule\noalign{}
Population & R squared & p value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textit{S. a. aberti Carson-SFW} & 0 & 0.962 \\
\textit{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
\textit{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
\textit{S. a. aberti San Juan} & 0 & 0.952 \\
\textit{S. a. barberi} & 0.794 & 0.001 \\
\textit{S. a. chuscensis E} & 0.002 & 0.892 \\
\textit{S. a. chuscensis W} & 0.018 & 0.713 \\
\textit{S. a. ferreus Carson E} & 0.04 & 0.581 \\
\textit{S. a. ferreus Pike} & 0 & 0.967 \\
\textit{S. a. kaibabensis} & 0.005 & 0.841 \\
\end{longtable}

@@

With the current availability of occurrence data and models of potential
distribution projected to past climate scenarios it is possible to find
geographic patterns of differentiation in populations at an acceptable
resolution (half a degree) that allows us to define sites where the
species can potentially be found but share different degrees of genetic
diversity.

Projections of genetic diversity in geography have been made in previous
studies Van Zonneveld et al. (2012). However in a general way they are
extrapolations of observations, without a predictive model, which with
some explanatory variable behind can explain in a statistically
convincing way the information obtained from population genetics.

@@

Thanks to the theoretical support that we have briefly outlined in this
paper, it is possible to explain the explanatory value of the
Coefficient of determination for our linear model (\$R\^{}2 = 0.79 \$),
which is information coming from two completely different sources,
without any correlation, but always guided by the premise of the
evolutionary process as causal.

@@

The generalization of our protocol and its implementation in tools
widely used by biogeography allow its direct implementation in many case
studies with available information, as well as the design of new
questions, e.g.~species assemblages under prevalence of environmental
suitability. A research path was also established on whether the process
of environmental stability over time is related to biodiversity.

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-owens2013constraints}{}}%
Owens, Hannah L, Lindsay P Campbell, L Lynnette Dornak, Erin E Saupe,
Narayani Barve, Jorge Soberón, Kate Ingenloff, et al. 2013.
{``Constraints on Interpretation of Ecological Niche Models by Limited
Environmental Ranges on Calibration Areas.''} \emph{Ecological
Modelling} 263: 10--18.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Environmental suitability throughout the late Quaternary explains population genetic diversity.},
  pdfauthor={Norma Hernandez, Angel Robles Nathan Upham},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Environmental suitability throughout the late Quaternary explains
population genetic diversity.}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

Genetic variation among populations is known to exhibit biogeographic
patterns in many species, but general rules of spatial genetic variation
have not been established. In this paper, we establish a theoretical
framework based on projecting environmental Grinellian niches back
through time to relate the present geographic distribution of population
genetic structure to a given species' historical evolutionary context.

Thanks to the advance of next-generation sequencing technologies, as
well as more accurate climate models and an enormous amount of
information stored in biological collections, it is possible to
implement this theoretical framework directly. We use as a case study
the Tassel-eared Squirrel (\emph{Sciurus aberti}) to jointly analyze
spatial, environmental, and genetic data to predict the historical
endemic area of this species. Our results reveal that in cases of
genetic isolation by geographic distance, the prevalence of
environmental suitability over time corresponds to the genetic fixation
index of populations with respect to a source population. Populations
closer to the historical endemic area present a higher genetic diversity
and a lower\(F_{st}\) value.

This empirical example fits into a theoretical framework, which allows
us two advances: (i) we can now add a biogeographic explanation to the
results obtained with population genetic methods; and (ii) we can
generate maps of this genetic structure as predictive tools to support
biodiversity conservation efforts. Overall, this work advances a
perspective that integrates population genetics with historical patterns
of species distribution.

@@

Current species distributions do not always correspond to their
historical distributions over evolutionarily significant periods. Often
they differ substantially, such as in the case of postglacial range
expansion in high latitude environments (e.g., Lessa, Cook, and Patton
(2003)). This is because environmental conditions are highly dynamic
over time, and species tend to track those environments, distributing
where conditions are most favorable. That is, species environmental
niches tend to be conserved (Peterson (2011)).

@@

With this approach we expect to predict the geographic pattern of
population-genetic structure from abiotic environmental information
alone. This approach is strongly driven by advances in currently
available climate simulations (Leonardi et al. 2023; Krapp et al. 2021),
as well as next-generation sequencing data, and supported by both
ecological niche (Thorup et al. 2021; Nogués-Bravo 2009; Lira-Noriega
and Manthey 2014) and population genetics theories (Walsh and Lynch
2018). Importantly, these predicted patterns of population structure
will serve as a type of \textit{a priori} hypothesis for the influence
of abiotic factors alone, with the empirical detection of deviations
from this hypothesis indicating that biotic factors such as competition
or predation from other species have also impacted the observed
population structure.

In this work, we propose a method to find an index that we call the
Suitability Prevalence Area (SPA) that is the area of highest
environmental suitability for a given species over a historical time
period. The SPA is an index with a double purpose: 1) identifying areas
of greatest likelihood for the historical distribution of a target
species (i.e., historical endemism); and 2) to explain the expected
patterns of genetic diversity given the species' inferred history of
population stability and connectivity.

As a case study, we consider Tassel-eared Squirrels
(\textit{Sciurus aberti}), a species of rodent that is currently
distributed in disjunct patches from the southern Rocky Mountains in the
United States to the northern Sierra Madre Occidental in Mexico for
which it is possible to find reliable information on both the genetic
structure of its populations and its current distribution (Bono et al.
2018; Burgin et al. 2018). To obtain the SPA for \textit{S. aberti}, we
overlapped historical reconstructions of the species' geographical
distribution estimated every 2,000 years from the present day until
120,000 years ago, recording the environmental suitability in each time
interval at each modeled site.

Subsequently, we calculated the fixation index (\(F_{st}\)) of the
genetic relatedness among populations and compared that to the SPA index
of historical endemism, constructing a statistical model of the fixation
index as a function of the SPA. With this explanatory model, the
\(F_{st}\) values were projected to the current geographic distribution
of \textit{S. aberti} to obtain a predictive map of expected \(F_{st}\)
values over space.

@@

In other words, \(\mathbf{G}_o\) is the set of sites in the geography
that present positive growth rates related to favorable conditions. If
we take the set of places in environmental space where these values are
greater than a threshold, we can define a volume in multidimensional
space such that \(R(\vec{e})\) is always positive and greater than this
threshold, that is
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

@@

Locally, the growth rate will be related to the local population size as
long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic
growth rate, which depends on the change in environmental conditions
over time, can be expressed as: \(N_i(t+1) = R(\vec{e_i}(t) ) N_i(t)\).
That is, the locally population size is a function of ideal
environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > 1 \}\) and therefore any
local population will have a positive growth rate.

@@

This equation was proposed by Soberon and Peterson (2020) to relate
demographics to a function of evolutionary fitness and environment. In
that study they tested the validity with monthly variables and empirical
values on a short time scale. relating life history parameters and the
environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate
probability density function, it would be equivalent to the distance
from the maximum of the function. That is, the fitness-related
environmental suitability can be any function such that it is \(0\) if
\(R_i(\vec{e}_i) > 1\) and \(1\) when \(R\) is evaluated at the expected
value of \(\vec {e}\) for all values of \(R_i(\vec{e}_i) > 1\). In a
practical way, \(\mathbf{N}_F\) is approximated from points of presence
in the geography obtained from databases, numerically approximating a
convex surface with a surface \(\mathbf{N}_R\) of the environmental
space.

@@

The goal of this theoretical approach is not to know the shape of the
function of \(R\) but rather to find this local variation relating this
value of \(R\) and the function \(\mathbf{N}_F\) with a function \$
S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be estimated
from known methods of applied mathematics and machine learning (for
example the case of Maxent, (Phillips, Anderson, and Schapire 2006)),
i.e.~:

\[
N_i \propto  S_i(\vec{e}_i)
\] Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) in
some point of the environmental space.

@@

On the other hand, loss of heterozygosity may be related to the amount
of genetic variation present in the absence of mutation and selection,
so theoretically, one would expect a strong correlation between
effective population size and heterozygosity on the basis of population
genetics. We start from the assumption that in the absence of gene flow
between subpopulations, the rate of fixation by isolation depends on the
effective population size \(N_e\) when subpopulations diverge from a
common ancestor in \(t\) generations. Walsh and Lynch (2018):

@@

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
We substitute the harmonic mean into the above equation, considering
that although the number of generations is not exactly equal to the time
of the fixation index, they are expected to be very close:

@@

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
where the \(F_{st}\) is lower are in areas where the prevalence over
time is higher, i.e.\textasciitilde environmental conditions remain
suitable for a significant period of time and therefore no effects of
genetic drift of populations have occurred. Conversely, it is expected
that in populations with a high fixation index value, suitability will
not have prevailed over the same time interval even though conditions
currently exist and populations are present.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past analogous climates and retain projections only for
  sites with climates analogous to the present.
\item
  Generate a harmonic mean suitability map for all time slices.
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  Obtain georeferenced points of genetically structured populations with
  fixation index values.
\item
  Extract values from the SPA map with these points.
\item
  Perform a linear model of the fixation index value with respect to the
  SPA.
\item
  With the linear model all SPA values of suitability are interpolated.
\item
  Generate a map with all the values interpolated by the model.
\end{itemize}

As a case study we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018). In order to select this species,
we list the following considerations:

@@

@@

\hypertarget{fixation-index-as-a-function-of-spa}{%
\subsubsection{Fixation index as a function of
SPA}\label{fixation-index-as-a-function-of-spa}}

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

\begin{longtable}[]{@{}ccc@{}}
\toprule\noalign{}
Population & R squared & p value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{S. a. aberti Carson-SFW} & 0 & 0.962 \\
\emph{S. a. aberti Coconino-Gila} & 0.131 & 0.304 \\
\emph{S. a. aberti MT-Zuni} & 0.002 & 0.894 \\
\emph{S. a. aberti San Juan} & 0 & 0.952 \\
\emph{S. a. barberi} & 0.794 & 0.001 \\
\emph{S. a. chuscensis E} & 0.002 & 0.892 \\
\emph{S. a. chuscensis W} & 0.018 & 0.713 \\
\emph{S. a. ferreus Carson E} & 0.04 & 0.581 \\
\emph{S. a. ferreus Pike} & 0 & 0.967 \\
\emph{S. a. kaibabensis} & 0.005 & 0.841 \\
\end{longtable}

@@

With the current availability of occurrence data and models of potential
distribution projected to past climate scenarios it is possible to find
geographic patterns of differentiation in populations at an acceptable
resolution (half a degree) that allows us to define sites where the
species can potentially be found but share different degrees of genetic
diversity..

Projections of genetic diversity in geography have been made in previous
studies (Van Zonneveld et al. 2012). However in a general way they are
extrapolations of observations, without a predictive model, which with
some explanatory variable behind can explain in a statistically
convincing way the information obtained from population genetics.

@@

Thanks to the theoretical support that we have briefly outlined in this
paper, it is possible to explain the explanatory value of the
coefficient of determination for our linear model, \(R^{2} = 0.79\),
which is information coming from two completely different sources, but
always guided by the premise of the evolutionary process as causal.

@@

The generalization of our protocol and its implementation in tools
widely used by biogeography allow its direct implementation in many case
studies with available information, as well as the design of new
questions, e.g.\textasciitilde species assemblages under prevalence of
environmental suitability. A research path was also established on
whether the process of environmental stability over time is related to
biodiversity.

It is also noteworthy to mention that in previous taxonomic
descriptions, it was taken into account that the distribution of Sciurus
aberti was only related to the distribution of Arizona pinus ponderosa.
However, there is a consensus on the range of distribution of this pine,
which is limited to the southern United States (Hess and Fulé 2020).
Therefore the known distribution of Sciurus aberti is in mixed
coniferous forests. This plasticity in the squirrel-pine interaction
explains the expansion of the species in interglacial periods and its
distribution in warmer periods, such as the current one. However, the
low genetic diversity found in places where the squirrel species
interact with few pine species could put it at greater risk of climate
change that constricts the distribution of p.~p.~Arizona, especially in
the Sky Island region of Arizona.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hess2020mexican}{}}%
Hess, Victoria A, and Peter Z Fulé. 2020. {``Is a Mexican Pine Species
Better Adapted to the Warming Climate of the Southwestern USA?''}
\emph{Frontiers in Forests and Global Change} 3: 60.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lessa2003genetic}{}}%
Lessa, Enrique P, Joseph A Cook, and James L Patton. 2003. {``Genetic
Footprints of Demographic Expansion in North America, but Not Amazonia,
During the Late Quaternary.''} \emph{Proceedings of the National Academy
of Sciences} 100 (18): 10331--34.

@@

\leavevmode\vadjust pre{\hypertarget{ref-peterson2011ecological}{}}%
Peterson, A Townsend. 2011. {``Ecological Niche Conservatism: A
Time-Structured Review of Evidence.''} \emph{Journal of Biogeography} 38
(5): 817--27.
